---
title: "Jobs Report Scheduled and Excel and Highcharter"
resource_files: config.yml
rmd_output_metadata:
  rsc_email_suppress_scheduled: yes
output:
  html_document:
    df_print: paged
  html_notebook: default
---



```{r setup, include = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = NA)
library(tidyverse)
library(tidyquant)
library(highcharter)
library(blastula)
library(formattable)
library(openxlsx)
library(config)
library(dbplyr)
library(DBI)
library(odbc)
library(htmltools)

dw <- config::get("mssql")

con <- DBI::dbConnect(odbc::odbc(),
                      Driver = dw$Driver,
                      Server = dw$Server,
                      #Host = dw$Host,
                      UID    = dw$UID,
                      PWD    = dw$PWD,
                      Port   = dw$Port,
                      Database = dw$Database
                      #Schema = dw$Schema
                      )

```

### Import data from sql

```{r}
bls_labor_for_data <-
  tbl(con, "bls_labor_for_data") %>% 
  collect() %>% 
  mutate(date = ymd(date),
         value = as.numeric(value))

fred_empl_data <-
  tbl(con, "fred_empl_data") %>% 
  collect()
```


### Chop up the Fred data into sectors and hours/wages objects

```{r}

labor_sector_data <- 
  fred_empl_data  %>% 
  filter(date >= "1990-01-01") %>%
  dplyr::select(1, 9:22) %>% 
  gather(sector, employees, -date) %>% 
  group_by(sector) %>% 
  mutate(monthly_change = employees - lag(employees, 1),
         yearly_change = employees - lag(employees, 12)) %>%
  rename(jobs = employees) %>% 
  na.omit()

hours_wages_data <-
  fred_empl_data%>%
  dplyr::select(date, `Avg Hourly Earnings`, `Avg Weekly Hours`)  %>% 
  na.omit()

```

### Visualize Total Employment Since 2009 with ggplot

```{r}
labor_sector_data_ggplot <- 
  labor_sector_data %>% 
  filter(sector == "Nonfarm Employment") %>% 
  filter(date >= "2009-01-01") %>% 
   mutate(col_blue = 
           if_else(monthly_change > 0, 
                  monthly_change, as.numeric(NA)),
         col_red = 
           if_else(monthly_change < 0, 
                  monthly_change, as.numeric(NA))) %>%
  ggplot(aes(x = date)) +
  geom_col(aes(y = col_red),
               alpha = .85, 
               fill = "pink", 
               color = "pink") +
  geom_col(aes(y = col_blue),
               alpha = .85, 
               fill = "cornflowerblue", 
               color = "cornflowerblue") +
  ylab("Monthly Change (thousands)") +
  labs(title = "Monthly Private Employment Change",
       subtitle = "total empl, since 2009",
       caption = "enjoy!") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption=element_text(hjust=0))

labor_sector_data_ggplot
```

### Visualize Total Employment Since 2009 with highcharter


```{r}
total_employ_hc <- 
  labor_sector_data %>% 
  filter(sector == "Nonfarm Employment") %>% 
  filter(date >= "2009-01-01") %>%
  mutate(coloract = ifelse(monthly_change > 0, "#6495ed", "#ffe6ea"))
```


```{r}
total_employ_hc_vis <- 
highchart() %>%
  hc_title(text = "Monthly Employment Change") %>%
  hc_add_series(total_employ_hc,  
                "column", 
                pointWidth = 5,
                hcaes(x = date,
                      y = monthly_change,
                      color = coloract),
                name = "monthly empl change") %>% 
  hc_xAxis(type = "datetime") %>% 
  hc_exporting(enabled = TRUE)

total_employ_hc_vis

```

```{r, echo = FALSE}
total_employ_hc_fname <- 
  paste("total_employ_hc_",Sys.Date(), ".xlsx", sep = "")

write.xlsx(total_employ_hc, file = total_employ_hc_fname)

rmarkdown::output_metadata$set(rsc_output_files = list(total_employ_hc_fname))
```

[Link to total nonfarm employment data Excel](`r total_employ_hc_fname`)

### Visualize Sector by Sector Change Last Month with ggplot


```{r}
recent_month_sector_ggplot <- 
labor_sector_data  %>% 
  filter(date == (last(date))) %>% 
   mutate(col_blue = 
           if_else(monthly_change > 0, 
                  monthly_change, as.numeric(NA)),
         col_red = 
           if_else(monthly_change < 0, 
                  monthly_change, as.numeric(NA))) %>%
  ggplot(aes(x = reorder(sector, monthly_change))) +
  geom_col(aes(y = col_red),
               alpha = .85, 
               fill = "pink", 
               color = "pink") +
  geom_col(aes(y = col_blue),
               alpha = .85, 
               fill = "cornflowerblue", 
               color = "cornflowerblue") +
  labs(title = paste((lubridate::month(last(labor_sector_data$date), 
                                       label = TRUE, abbr = FALSE)), 
                     "Employment Change", 
                     sep = " "),
       subtitle = "by sector, and total", 
       x = "sector", 
       y = "Monthly Change (thousands)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

recent_month_sector_ggplot
```

### Visualize Sector by Sector Change Last Month with highcharter

```{r}
month <- lubridate::month(last(labor_sector_data$date),
                                 label = TRUE, 
                                 abbr = FALSE)
  
labor_sector_most_recent_month <- 
  labor_sector_data  %>% 
  filter(date == (last(date))) %>%
  arrange(monthly_change) %>% 
  mutate(coloract = if_else(monthly_change > 0, "#6495ed", "#ffe6ea"))


highchart() %>%
  hc_title(text = paste(month, "Employment Change", sep = " ")) %>%
  hc_add_series(labor_sector_most_recent_month,  
                "column", 
                pointWidth = 20,
                hcaes(x = sector,
                      y = monthly_change,
                      color = coloract),
                showInLegend = FALSE) %>% 
  hc_xAxis(categories = labor_sector_most_recent_month$sector) %>%
  hc_xAxis(title = list(text = "Sector")) %>% 
  hc_yAxis(title = list(text = "Monthly Change (thousands)"))


```

```{r, echo = FALSE}
labor_sector_most_recent_month_fname <- 
  paste("labor_sector_most_recent_month_", Sys.Date(), ".xlsx", sep = "")

write.xlsx(labor_sector_most_recent_month, file = labor_sector_most_recent_month_fname)

rmarkdown::output_metadata$set(rsc_output_files = list(labor_sector_most_recent_month_fname))
```

[Link to sector change last month data Excel](`r labor_sector_most_recent_month_fname`)

### Visualize Sector by Sector Data over time with ggplot

```{r, fig.height=10, fig.width=12}

labor_sector_data %>% 
  filter(sector != "Nonfarm Employment") %>% 
  filter(date >= (last(date) - months(24))) %>% 
   mutate(col_blue = 
           if_else(monthly_change > 0, 
                  monthly_change, as.numeric(NA)),
         col_red = 
           if_else(monthly_change < 0, 
                  monthly_change, as.numeric(NA))) %>%
  ggplot(aes(x = date)) +
  geom_col(aes(y = col_red),
               alpha = .85, 
               fill = "pink", 
               color = "pink") +
  geom_col(aes(y = col_blue),
               alpha = .85, 
               fill = "cornflowerblue", 
               color = "cornflowerblue") +
  ylab("Monthly Change (thousands)") +
  labs(title = "Monthly Private Employment Change",
       subtitle = "by sector, 24 months") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
    strip.background = element_blank(),
    strip.placement = "inside",
    strip.text = element_text(size=15),
    panel.spacing = unit(0.2, "lines") ,
    panel.background=element_rect(fill="white")) +
  facet_wrap(~sector, shrink = FALSE)
```


### Visualize Sector by Sector Data over time with highcharter

```{r}
labor_sector_data_spread_hc <- 
  labor_sector_data %>% 
  filter(sector != "Nonfarm Employment") %>% 
  filter(date >= (last(date) - months(24))) %>% 
  select(date, sector, monthly_change) %>% 
  spread(sector, monthly_change)

map(names(labor_sector_data_spread_hc[2:11]), function(x){

labor_data_hc_fun <- 
  labor_sector_data %>% 
  filter(sector == x) %>% 
  filter(date >= (last(date) - months(24))) %>% 
   mutate(coloract = if_else(monthly_change > 0, "#6495ed", "#ff9999"))

highchart() %>%
  hc_title(text = paste(x, "Monthly Employment Change", sep = " ")) %>%
  hc_add_series(labor_data_hc_fun,  
                "column", 
                pointWidth = 10,
                hcaes(x = date,
                      y = monthly_change,
                      color = coloract),
                name = "monthly empl change") %>% 
  hc_xAxis(type = "datetime") %>% 
  hc_legend(enabled = FALSE) %>% 
  hc_exporting(enabled = TRUE)
}) %>% 
  hw_grid(rowheight = 300, ncol = 3) %>% 
  browsable()
```

```{r, echo = FALSE}
labor_sector_fname <- 
  paste("labor_sector_",Sys.Date(), ".xlsx", sep = "")

write.xlsx(labor_sector_data_spread_hc, file = labor_sector_fname)

rmarkdown::output_metadata$set(rsc_output_files = list(labor_sector_fname))
```

[Link to sector by sector labor data Excel](`r labor_sector_fname`)


### Grab the ADP and BLS Data from the Fred object

```{r}
adp_fred <-
  fred_empl_data  %>%
  select(date, ADP = `ADP Estimate`, BLS = `Nonfarm Employment`)  %>% 
    gather(report, month, -date) %>% 
    group_by(report) %>% 
    mutate(monthly_change = month - lag(month, 1)) %>% 
  filter(date >= "2017-01-01") %>% 
  select(-month)
```

```{r, echo = FALSE}
adp_fred_fname <- 
  paste("adp_fred_",Sys.Date(), ".xlsx", sep = "")

write.xlsx(adp_fred, file = adp_fred_fname)

rmarkdown::output_metadata$set(rsc_output_files = list(adp_fred_fname))
```

[Link to ADP and BLS data Excel](`r adp_fred_fname`)

### ADP v. BLS ggplot

```{r}

adp_fred %>%
  ggplot(aes(x = date, y = monthly_change, fill = report)) + 
  geom_col(position = "dodge") + 
  scale_fill_manual(values = c("pink","cornflowerblue")) +
  labs(title = "ADP Estimate v. BLS Report",
       caption = "data source: fred",
       y = "monthly change (thousands)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        plot.caption  = element_text(hjust = 0)) +
  scale_x_date(breaks = scales::pretty_breaks(n = 16))
```

### ADP v. BLS highcharter

```{r}
adp_bls_hc <- 
  adp_fred %>% 
  mutate(coloract = ifelse(report == "ADP", "#FF0000", "#00FFFF"))

hchart(adp_bls_hc, 
       type = 'column', 
       hcaes(y = monthly_change,
             x = date,
             group = report, 
             color = coloract)) %>% 
  hc_title(text = "ADP v. BLS") %>%
  hc_xAxis(title = list(text = "")) %>%
  hc_yAxis(title = list(text = "monthly change (thousands)")) %>%
  hc_add_theme(hc_theme_flat()) %>%
  hc_exporting(enabled = TRUE) %>% 
  hc_legend(enabled = FALSE)
```

### Vis from BLS direct download

```{r}
recessions_tribble <- tribble(
  ~Peak, ~Trough,
  "1948-11-01", "1949-10-01",
  "1953-07-01", "1954-05-01",
  "1957-08-01", "1958-04-01",
  "1960-04-01", "1961-02-01",
  "1969-12-01", "1970-11-01",
  "1973-11-01", "1975-03-01",
  "1980-01-01", "1980-07-01",
  "1981-07-01", "1982-11-01",
  "1990-07-01", "1991-03-01",
  "2001-03-01", "2001-11-01",
  "2007-12-01", "2009-06-01") %>% 
  mutate(Peak = ymd(Peak),
         Trough = ymd(Trough))

# can save ggplot variables
# geom_rect creates that object
recession_shade <- 
  geom_rect(data = recessions_tribble, 
            inherit.aes = F, 
            aes(xmin = Peak, 
                xmax = Trough, 
                ymin = -Inf, 
                ymax = +Inf), 
            fill = 'darkgray', 
            alpha = 0.4)

bls_labor_for_data %>% 
  group_by(dem) %>%
  filter(date >= "1980-01-01") %>% 
  ggplot(aes(x = date, y = value, color = dem, label = dem)) +
  recession_shade +
  geom_line()  +
  theme_minimal() +
  scale_x_date(date_breaks = "2 years", 
               date_labels = "%Y") +
  xlim(ymd("1980-01-01"), Sys.Date()) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10), 
                     labels = function(x) paste0(x, "%"), 
                     sec.axis = dup_axis()) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5),
        plot.caption=element_text(hjust=0)) +
  ylab("") +
  xlab("") +
  ggtitle("Labor Force Participation Rate: Prime Working Age (25-54)", 
          subtitle = "seasonally adjusted") +
  labs(caption = "based on work by the inestimable @lenkiefer") 
```


```{r}
library(timetk)
bls_labor_for_data_xts <- 
  bls_labor_for_data %>% 
  group_by(dem) %>%
  filter(date >= "1980-01-01") %>% 
  spread(dem, value) %>% 
  tk_xts(date_var = date)


highchart(type = 'stock')%>%
  hc_add_series(bls_labor_for_data_xts$All, name = "All") %>%
  hc_add_series(bls_labor_for_data_xts$Men, name = "Men") %>%
  hc_add_series(bls_labor_for_data_xts$Women, name = "Women") %>%
  hc_title(text = "Labor Force Participation") %>% 
      hc_xAxis(type = 'datetime',
                 plotBands = list(
                   list(
                     label = list(text = "Bush Senior"),
                     color = "rgba(100, 0, 0, 0.1)",
                     from = datetime_to_timestamp(as.Date('1990-07-01', tz = 'UTC')),
                     to = datetime_to_timestamp(as.Date('1991-03-01', tz = 'UTC'))
                     ),
                   list(
                     label = list(text = "Internet Bubble"),
                     color = "rgba(100, 0, 0, 0.1)",
                     from = datetime_to_timestamp(as.Date('2001-03-01', tz = 'UTC')),
                     to = datetime_to_timestamp(as.Date('2001-11-01', tz = 'UTC'))
                     ),
                   list(
                     label = list(text = "Financial Crisis"),
                     color = "rgba(100, 0, 0, 0.1)",
                     from = datetime_to_timestamp(as.Date('2007-12-01', tz = 'UTC')),
                     to = datetime_to_timestamp(as.Date('2009-06-01', tz = 'UTC'))
                     ),
                   list(
                     label = list(text = "Volker Rate Rise"),
                     color = "rgba(100, 0, 0, 0.1)",
                     from = datetime_to_timestamp(as.Date('1981-07-01', tz = 'UTC')),
                     to = datetime_to_timestamp(as.Date('1982-11-01', tz = 'UTC'))
                     )
                   )
               ) %>% 
  hc_navigator(enabled = FALSE) %>% 
  hc_scrollbar(enabled = FALSE) %>% 
  hc_legend(enabled = TRUE) %>% 
  hc_exporting(enabled = TRUE)

```

```{r, echo = FALSE}
labor_for_part_fname <- 
  paste("labor_for_part_",Sys.Date(), ".xlsx", sep = "")

write.xlsx(bls_labor_for_data_xts, file = labor_for_part_fname)

rmarkdown::output_metadata$set(rsc_output_files = list(labor_for_part_fname))

```

[Link to labor force part data Excel](`r labor_for_part_fname`)


### Schedule email for BLS release dates

```{r}
# we will schedule this to run every friday in the RSC scheduler.
# On one Friday per month, when BLS releases jobs data, 
# this report will be emailed. 

release_dates <- 
tribble(
~"bls release date",  
"Jan. 05, 2018",
"Feb. 02, 2018",
"Mar. 09, 2018",
"Apr. 06, 2018",
"May. 04, 2018",
"Jun. 02, 2018",
"Jul. 06, 2018",
"Aug. 03, 2018",
"Sep. 07, 2018",
"Oct. 05, 2018",
"Nov. 02, 2018",
"Dec. 07, 2018") %>% 
mutate(`bls release date` = 
         ymd(parse_date_time(`bls release date`, "%m%d%Y"))) %>%
pull(`bls release date`)

 
if (today() %in% release_dates) {
  # send email if today is a release today
  rmarkdown::output_metadata$set(rsc_email_suppress_scheduled = FALSE)
  
  # embed charts and data table
  
  sign_formatter <- formatter("span", 
  style = x ~ style(color = ifelse(x > 0, "green", 
    ifelse(x < 0, "red", "black"))))


tbl <- 
  labor_sector_most_recent_month %>% 
  dplyr::select(-coloract) %>% 
  format_table(x = .,
               list(monthly_change = sign_formatter,
                    yearly_change = sign_formatter,
                    jobs = comma))
last_month <- lubridate::month(last(labor_sector_data$date), 
                                       label = TRUE, abbr = FALSE)
  
  msg <- compose_email(
    body = "
    Hello Team,<br/>
    
    Here are the latest jobs report data and charts,

    updated with the BLS numbers for {last_month}.<br/>
  
    <br/>Here is monthly change over time: 

    {add_ggplot(labor_sector_data_ggplot, width = 6, height = 6)}


    Here are sector by sector jobs numbers for {last_month}: 

    {add_ggplot(recent_month_sector_ggplot, width = 6, height = 6)}

    Here are the raw numbers for {last_month}:

    {tbl}
    

    Let me know if you have any questions.
    

    Arrividerci,
    JKR"
  )
  rmarkdown::output_metadata$set(rsc_email_body_html = msg$html_str)
  rmarkdown::output_metadata$set(rsc_email_images = msg$images)
  
  # attach the excel file
  rmarkdown::output_metadata$set(rsc_email_attachments = list( total_employ_hc_fname, labor_sector_most_recent_month_fname))
  
} else {
  # don't send an email if today is not a release date
  rmarkdown::output_metadata$set(rsc_email_suppress_scheduled = TRUE)
}



```


